Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-26 2024pipit/section/50-Extraction.tex
extraction, evaluation

  \begin{tabbing}
  MM \= update: \= = \= @let (sum,y) @ \= \kill
  @let@ sum_def: system (\textit{ints}: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0; \}; \\
  \> step  \> = $\lambda{} i~s.$ \\
  \> \> \> @let@ $(\textit{fby}_0, s_0)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \bot_\ZZ \})$ @in@ \\
  \> \> \> @let@ $(\textit{sum}_0, s_0)$ \> $= (\textit{fby}_0 + i.\text{ints}, s_0)$ @in@ \\
  \> \> \> @let@ $(\textit{fby}_1, s_1)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \textit{sum}_0 \})$ @in@ \\
  \> \> \> @let@ $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ @in@ \\
  \> \> \> $(\textit{sum}_0, s_1)$
  \end{tabbing}

@@

% The generated C code includes a struct containing the

  \begin{tabbing}
    @struct sum_state { int sum_fby; }@ \\
    @void   sum_init(struct sum_state* state);@ \\
    @int    sum_step(struct sum_state* state, int ints);@\\
  \end{tabbing}

>>>>

  \begin{tabbing}
  MM \= update: \= = \= @let (sum,y) @ \= \kill
  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0; \}; \\
  \> step  \> = $\lambda{} i~s.$ \\
  \> \> \> @let@ $(\textit{fby}_0, s_0)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \bot_\ZZ \})$ @in@ \\
  \> \> \> @let@ $(\textit{sum}_0, s_0)$ \> $= (\textit{fby}_0 + i.\text{ints}, s_0)$ @in@ \\
  \> \> \> @let@ $(\textit{fby}_1, s_1)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \textit{sum}_0 \})$ @in@ \\
  \> \> \> @let@ $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ @in@ \\
  \> \> \> $(\textit{sum}_0, s_1)$ \}
  \end{tabbing}

@@

The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions; this interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular}:
  \begin{tabbing}
    @struct sum_state { uint32_t sum_fby; }@ \\
    @void   sum_reset(struct sum_state* state);@ \\
    @int    sum_step(struct sum_state* state, uint32_t ints);@
  \end{tabbing}

The reset function takes the pointer to the state struct and sets it to its initial values.
The step function takes the pointer to the state struct and the inputs, and returns the result integer.
The state struct is updated in-place.
The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.

### 2024-03-26 2024pipit/section/60-Evaluation.tex
todos
extraction, evaluation

The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
We generate real-time C code as described in \autoref{s:extraction}.
We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent.
For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$.
Further evaluation is required to ensure that the complete runtime including the hardware-interop layer is sufficient for full-speed CAN.

@@

We were able to reliably verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, we ran into intermittent runtime errors.
% Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after several hours and multiple runtime errors.
For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.

We made a critical simplification in the Kind2 implementation, which was to modify the trigger-enabled set to be a single cycle index.
In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
In our Pipit development, we can treat the definition of the cycle set abstractly.
However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
This limitation also makes the specification quite unwieldy, as functions must be manually inlined.

@@

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & user-time & wall-clock & user-time & wall-clock & user-time \\
  \hline

  1 & 1.51s&1.33s
  & 1.61s&2.26s
  & 5.28s&5.03s \\
  2 & 1.50s&1.29s
  & 1.68s&2.85s
  & 5.28s&5.03s \\
  4 & 1.56s&1.59s
  & 2.10s&5.05s
  & 5.28s&5.03s \\
  8 & 1.80s&3.12s
  & 4.29s&18.12s
  & 5.28s&5.03s \\
  16 & 3.41s&12.29s
  & 16.41s&71.23s
  & 5.28s&5.03s \\
  32 & 11.24s&67.27s
  & 941.64s&3853.16s
  & 5.28s&5.03s \\
\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification. The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array. We were unable to verify arrays of size 64 with Kind2 as our verification timed out after three hours.}
\label{f:evaluation:kind2-runtime}
\end{figure}

>>>>

The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
We generate real-time C code as described in \autoref{s:extraction}.
We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $140\mu{}s$, while the average was $90\mu{}s$ with a standard deviation of $1.5\mu{}s$.
Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent.
For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$.
Further evaluation is required to ensure that the complete runtime including the hardware-interop layer is sufficient for full-speed CAN.

@@

We were able to encode equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
For example, the specification-only function that finds the next trigger is naturally recursive.
Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
For example, given a hypothetical recursive definition of @next@ (left), the array-scan encoding is (right):

\begin{minipage}{0.4\textwidth}
\begin{verbatim}
next(i) =
  if enabled(triggers, i, c)
  then i
  else if i >= LENGTH - 1
  then NO_NEXT_TRIGGER
  else next (i + 1);
\end{verbatim}
\end{minipage}
\begin{minipage}{0.6\textwidth}
\begin{verbatim}
next_array[i] =
  if enabled(triggers, LENGTH - 1 - i, c)
  then (LENGTH - 1 - i)
  else if i <= 0
  then NO_NEXT_TRIGGER
  else next_array[i - 1];
next(i) = next_array[LENGTH - 1 - i]

\end{verbatim}
\end{minipage}

% Ignoring bounds checks, this roughly corresponds to @next(i) = if enabled(i) then i else next(i + 1)@.
% However, Kind2 does not support recursive functions; instead, we were able to express the recursion using Kind2's array comprehension syntax, which allows one to define an array as a function of its index.
% This encoding was further complicated by the fact that array comprehensions can only depend on values of the array with lower indices, not ones with a higher index: concretely, we cannot write @next[i] = if enabled(i) then i else next[i + 1]@; instead, we invert the indices: @next[i] = if enabled(LENGTH - 1 - i) then LENGTH - 1 - i else next[i - 1]@.

We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
In our Pipit development, we can treat the definition of the cycle set abstractly.
However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
This limitation also makes the specification quite unwieldy, as functions must be manually inlined.

\TODO{include comparison of required proof size vs runnable code}

@@

We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 32 elements; above that, our verification timed out after one hour.
For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & CPU-time & wall-clock & CPU-time & wall-clock & CPU-time \\
  \hline

 1 & 1.48s&1.06s
 & 1.57s&2.26s
 & 5.25s&5.03s \\
 2 & 1.51s&1.26s
 & 1.71s&2.93s
 & 5.25s&5.03s \\
 4 & 1.57s&1.62s
 & 2.08s&4.78s
 & 5.25s&5.03s \\
 8 & 1.76s&3.07s
 & 4.21s&16.98s
 & 5.25s&5.03s \\
 16 & 3.36s&11.91s
 & 13.82s&65.57s
 & 5.25s&5.03s \\
 32 & 12.15s&62.38s
 & 269.14s&1230.05s
 & 5.25s&5.03s \\
 64 & 1701.01s&9096.99s
 &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\
 128 &  \multicolumn{2}{c|}{(timeout)}  &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\

% VERSION WITH SYSTEM TIME:
%  1 & 1.48s&1.06s&0.46s
%  & 1.57s&2.26s&0.84s
%  & 5.25s&5.03s&0.22s \\
%  2 & 1.51s&1.26s&0.40s
%  & 1.71s&2.93s&0.98s
%  & 5.25s&5.03s&0.22s \\
%  4 & 1.57s&1.62s&0.44s
%  & 2.08s&4.78s&1.03s
%  & 5.25s&5.03s&0.22s \\
%  8 & 1.76s&3.07s&0.51s
%  & 4.21s&16.98s&1.67s
%  & 5.25s&5.03s&0.22s \\
%  16 & 3.36s&11.91s&0.78s
%  & 13.82s&65.57s&3.71s
%  & 5.25s&5.03s&0.22s \\
%  32 & 12.15s&62.38s&2.43s
%  & 269.14s&1230.05s&30.93s
%  & 5.25s&5.03s&0.22s \\
%  64 & 1701.01s&9096.99s&41.88s
%  &  \multicolumn{2}{c|}{error}  & 5.25s&5.03s&0.22s \\
%  128 &  \multicolumn{2}{c|}{error}  &  \multicolumn{2}{c|}{error}  & 5.25s&5.03s&0.22s \\

\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
The ``wall-clock time'' denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
The time limit was one hour.}
\label{f:evaluation:kind2-runtime}
\end{figure}
